---
title: "Bellabeat Case Study (DA Google Certificate)"
author: "Pablo Viruell"
date: "2024-10-08"
output:
    github_document:
    df_print: paged
---

# Bellabeat Case Study (Google Data Analytics Certificate)

The following document shows the cleaning and analyzing steps made for Google's Bellabeat case study.

#### Step 1. Load the necessary packages

```{r include=FALSE}
library(tidyverse) #system of packages useful for data analysis
library(readr) #data import
library(tidyr) #data cleaning
library(janitor) #data manipulation
library (plyr) #data manipulation
library (dplyr) #data manipulation
library(lubridate) #datetime manipulation
library(ggplot2) #data visualizations
library(cowplot) #data visualization features
```

#### Step 2. Import the data

```{r eval=FALSE, include=FALSE}
#Setting a working directory
setwd("C:/Amryt/Data Analytics/Case study/BellaBeat case study/Fitabase Data 4.12.16-5.12.16")
#Importing the data 
daily_activity <- read.csv("dailyActivity_merged.csv")
hourly_calories <- read.csv("hourlyCalories_merged.csv")
hourly_steps <- read.csv("hourlySteps_merged.csv")
sleep <- read.csv("SleepDay_merged.csv")
weight <- read.csv("weightLogInfo_merged.csv")
```

#### Step 3. Clean the data

*3.1 Checking for NA values and duplicates*

```{r}
#Checking for NA values
sum(is.na(daily_activity))
sum(is.na(hourly_calories))
sum(is.na(hourly_steps))
sum(is.na(sleep))
sum(is.na(weight))

#Checking for duplicates
sum(duplicated(daily_activity))
sum(duplicated(hourly_calories))
sum(duplicated(hourly_steps))
sum(duplicated(sleep))
sum(duplicated(weight))

#Dropping the duplicate values found in table "sleep"
sleep <- unique(sleep)

#Verifying duplicate values were dropped
sum(duplicated(sleep))
```

*3.2 Checking data content and format*

```{r}
#Checking the number of unique users in each table
n_distinct(daily_activity$Id)
n_distinct(hourly_calories$Id)
n_distinct(hourly_steps$Id)
n_distinct(sleep$Id)
n_distinct(weight$Id)
```

Given the small amount of unique users in the table "weight", no reliable insights will be obtained. Therefore, it will be dropped from the analysis.

```{r}
#Reviewing the content of each remaining table
head(daily_activity)
head(hourly_calories)
head(hourly_steps)
head(sleep)

#Reviewing the format of each table
str(daily_activity)
str(hourly_calories)
str(hourly_steps)
str(sleep)
```

*3.3 Make necessary corrections to data format*

```{r}
#Renaming the table columns
daily_activity <- rename(daily_activity, Date= ActivityDate)
hourly_calories <- rename(hourly_calories, Datetime= ActivityHour)
hourly_steps <- rename(hourly_steps, Datetime= ActivityHour)
sleep <- rename(sleep, Date= SleepDay)

#Deleting the time section in the "sleep" table
sleep$Date <- gsub("12:00:00 AM", "", sleep$Date)

#Converting character format to date format
daily_activity$Date <- as_date(daily_activity$Date, format="%m/%d/%Y")
sleep$Date <- mdy(sleep$Date)

#Converting character format to POSIXct format
hourly_calories$Datetime <- convert_to_datetime(hourly_calories$Datetime, tz="UTC", character_fun = lubridate::mdy_hms)
hourly_steps$Datetime <- convert_to_datetime(hourly_steps$Datetime, tz="UTC", character_fun = lubridate::mdy_hms)

#Splitting the datetime columns
hourly_calories$Time <- format(as.POSIXct(hourly_calories$Datetime), format="%H:%M:%S")
hourly_calories$Date <- as.Date(hourly_calories$Datetime)
hourly_steps$Time <- format(as.POSIXct(hourly_steps$Datetime), format="%H:%M:%S")
hourly_steps$Date <- as.Date(hourly_steps$Datetime)

#Converting number format to character format
daily_activity$Id <- as.character(daily_activity$Id)
sleep$Id <- as.character(sleep$Id)
hourly_calories$Id <- as.character(hourly_calories$Id)
hourly_steps$Id <- as.character(hourly_steps$Id)
```

*3.4 Merge the data to facilitate its future use*

```{r}
#Merging the daily data
daily_data <- merge(daily_activity, sleep, by=c("Id", "Date"), all=TRUE)
```

*3.5 Create a new Weekday column, which will be useful for future analysis*

```{r}
daily_data <- daily_data %>%
  mutate(Weekday=wday(daily_data$Date, label=TRUE, abbr=FALSE, locale="UTC"))
```

#### Step 4. Analyze the data

*4.1 Basic Statistics*

```{r}
#Calculating basic statistics
summary(daily_data)
summary(sleep)
```

*4.2. Review and plot data by day of the week*

```{r}
#Number of data entries per weekday
ggplot(daily_data, aes(x=Weekday))+
  geom_bar(fill="orangered1")+
  labs(title="Data recorded per weekday",
       x=NULL,
       y="Number of data entries")+
  geom_text(aes(label=after_stat(count)),
            stat="count",
            size=3,
            position=position_stack(vjust=1.05))

#Average steps per weekday
daily_data %>%
  group_by(Weekday) %>%
  summarise(Mean_Steps=mean(TotalSteps)) %>%
  ggplot(aes(x=Weekday,
             y=Mean_Steps))+
  geom_col(fill="lightblue")+
  labs(title="Average steps taken per weekday",
       x=NULL,
       y="Average Steps")+
  geom_text(aes(label=ceiling(Mean_Steps)),
            size=3,
            hjust=0.5,
            position=position_stack(vjust=1.05))

#Minutes asleep per weekday
daily_data %>%
  group_by(Weekday) %>%
  summarise(Mean_Asleep=mean(TotalMinutesAsleep, na.rm=TRUE)) %>%
  ggplot(aes(x=Weekday,
             y=Mean_Asleep))+
  geom_col(fill="lightgreen")+
  labs(title="Average time asleep per weekday",
       x=NULL,
       y="Minutes Asleep")+
  geom_text(aes(label=ceiling(Mean_Asleep)),
            size=3,
            hjust=0.5,
            position=position_stack(vjust=1.05))
```

To plot the activity level per weekday, some data manipulation is needed first.

```{r}
#Create a new data frame with average intensity data
Intensity_mean <- daily_data %>%
  group_by(Weekday) %>%
  summarise(mean_very_active=mean(VeryActiveMinutes),
            mean_fairly_active=mean(FairlyActiveMinutes),
            mean_lightly_active=mean(LightlyActiveMinutes),
            mean_sedentary=mean(SedentaryMinutes))
Intensity_mean <- Intensity_mean %>%
  mutate_if(is.numeric, round)

#Reshape data frame from wide to long format
Intensity_mean_long <- pivot_longer(Intensity_mean,
                                    cols=c("VeryActiveMinutes","FairlyActiveMinutes", "LightlyActiveMinutes", "SedentaryMinutes"), 
                                    names_to = "Activity_Level",
                                    values_to = "Mean_Minutes")

#Calculate the percentage of minutes for each activity level per day
Intensity_mean_long=ddply(Intensity_mean_long, .(Weekday), transform, Ratio=Mean_Minutes/sum(Mean_Minutes)*100)
Intensity_mean_long$Daily_Percent=paste0(sprintf(Intensity_mean_long$Ratio), "%")
```

With the data ready, we can plot activity level per weekday.

```{r}
#Defining the order in which the activity level should appear in the plot
Intensity_mean_long$Activity_Level <- factor(Intensity_mean_long$Activity_Level,levels=c("mean_sedentary", "mean_lightly_active", "mean_fairly_active", "mean_very_active"))

#Plotting activity level per weekday
ggplot(Intensity_mean_long, aes(x=Weekday,
                                y=Mean_Minutes,
                                fill=Activity_Level))+
  geom_col()+
  labs(title="Activity Level per Weekday",
       x=NULL,
       y="Minutes")+
  guides(fill=guide_legend(title="Activity Level"))+
  scale_fill_discrete(labels=c("Very Active",
                               "Fairly Active",
                               "Lightly Active",
                               "Sedentary"))+
  geom_text(aes(label=ifelse(Mean_Minutes>100, Daily_Percent, '')), 
            size=3,
            hjust=0.5,
            position=position_stack(vjust=0.5))+
  geom_text(aes(label=after_stat(y), group=Weekday),
            stat="summary", fun=sum, vjust=1)
```

*4.3 Review and plot hourly data*

```{r}
#Average Hourly Steps
hourly_steps %>%
  group_by(Time) %>%
  summarise(Average_Steps=mean(StepTotal)) %>%
  ggplot(aes(x=Time,
             y=Average_Steps,
             fill=Average_Steps))+
  geom_col()+
  labs(title="Average Step Distribution per Hour",
       x=NULL,
       y="Average Steps",
       fill=NULL)+
  theme(axis.text.x=element_text(angle=90))+
  scale_fill_viridis_c()

#Average Hourly Calories
hourly_calories %>%
  group_by(Time) %>%
  summarise(Average_Calories=mean(Calories)) %>%
  ggplot(aes(x=Time,
             y=Average_Calories,
             fill=Average_Calories))+
  geom_col()+
  labs(title="Average Calorie Distribution per Hour",
       x=NULL,
       y="Average Calories",
       fill=NULL)+
  theme(axis.text.x=element_text(angle=90))+
  scale_fill_viridis_c()
```

*4.4 Review and plot individual data*

```{r}
#Calories vs Total steps
ggplot(daily_data, aes(x=TotalSteps,
                       y=Calories,
                       color=VeryActiveMinutes))+
  geom_point()+
  geom_smooth(method="loess")+
  scale_color_gradient(low="green", high="red")+
  labs(title="Calories vs Total Steps per Day",
       x="Total Steps",
       caption="Based on daily data")

#Activity Minutes vs Total steps for each activity level
ggplot(daily_data)+
  #Very Active Minutes
  geom_point(aes(x=TotalSteps,
                 y=VeryActiveMinutes),
             color="navy")+ 
  geom_smooth(method="loess", aes(x=TotalSteps,
                                  y=VeryActiveMinutes),
              color="navy")+
  #Fairly Active Minutes
  geom_point(aes(x=TotalSteps,
                 y=FairlyActiveMinutes),
             color="orange4")+
  geom_smooth(method="loess", aes(x=TotalSteps,
                                  y=FairlyActiveMinutes),
              color="orange4")+
  #Lightly Active Minutes
  geom_point(aes(x=TotalSteps,
                 y=LightlyActiveMinutes),
             color="chartreuse4")+
  geom_smooth(method="loess", aes(x=TotalSteps,
                                  y=LightlyActiveMinutes),
              color="chartreuse4")+
  #Sedentary Minutes
  geom_point(aes(x=TotalSteps,
                 y=SedentaryMinutes),
             color="deeppink4")+
  geom_smooth(method="loess", aes(x=TotalSteps,
                                  y=SedentaryMinutes),
              color="deeppink4")+
  labs(title="Activity Minutes vs Daily Steps",
       x="Total Daily Steps",
       y="Daily Activity Minutes")+
  annotate("text", x=30000,
           y=175,
           label="Very Active",
           fontface="italic",
           color="navy",
           size=3)+
  annotate("text", x=30000,
           y=70,
           label="Fairly Active",
           fontface="italic",
           color="orange4",
           size=3)+
  annotate("text", x=30000,
           y=275,
           label="Lightly Active",
           fontface="italic",
           color="chartreuse4",
           size=3)+
  annotate("text", x=30000,
           y=1075,
           label="Sedentary",
           fontface="italic",
           color="deeppink4",
           size=3)

#Activity Minutes vs Calories for each activity level
ggplot(daily_data)+
  #Very Active Minutes
  geom_point(aes(x=Calories,
                 y=VeryActiveMinutes),
             color="navy")+
  geom_smooth(method="loess", aes(x=Calories,
                                  y=VeryActiveMinutes),
              color="navy")+
  #Fairly Active Minutes
  geom_point(aes(x=Calories,
                 y=FairlyActiveMinutes),
             color="orange4")+
  geom_smooth(method="loess", aes(x=Calories,
                                  y=FairlyActiveMinutes),
              color="orange4")+
  #Lightly Active Minutes
  geom_point(aes(x=Calories,
                 y=LightlyActiveMinutes),
             color="chartreuse4")+
  geom_smooth(method="loess", aes(x=Calories,
                                  y=LightlyActiveMinutes),
              color="chartreuse4")+
  #Sedentary Minutes
  geom_point(aes(x=Calories,
                 y=SedentaryMinutes),
             color="deeppink4")+
  geom_smooth(method="loess", aes(x=Calories,
                                  y=SedentaryMinutes),
              color="deeppink4")+
  labs(title="Activity Minutes vs Calories",
       x="Daily Calories",
       y="Daily Activity Minutes")+
  annotate("text", x=4850,
           y=125,
           label="Very Active",
           fontface="italic",
           color="navy",
           size=3)+
  annotate("text", x=4850,
           y=-25,
           label="Fairly Active",
           fontface="italic",
           color="orange4",
           size=3)+
  annotate("text", x=4850,
           y=280,
           label="Lightly Active",
           fontface="italic",
           color="chartreuse4",
           size=3)+
  annotate("text", x=4850,
           y=480,
           label="Sedentary",
           fontface="italic",
           color="deeppink4",
           size=3)

#Time Asleep vs Sedentary Minutes
ggplot(daily_data)+
  geom_point(aes(x=TotalMinutesAsleep,
                 y=SedentaryMinutes))+
  geom_smooth(method="loess", aes(x=TotalMinutesAsleep,
                                  y=SedentaryMinutes))+
  labs(title="Sleep Time vs Sedentary Minutes",
       x="Daily Total Minutes Asleep",
       y="Sedentary Minutes",
       caption="Based on reported data. NA values not considered")+
  annotate("text", x=750,
           y=350,
           label="italic(R)^2==0.3597",
           parse=TRUE,
           size=3)
```

Based on the Time Asleep vs Sedentary Minutes plot, it seems like there might be a relationship between the variables. Therefore, we evaluate if true via a linear regression analysis.

```{r}
#Linear Regression analysis for Time Asleep vs Sedentary Minutes
Asleep_vs_Sedentary <- lm(TotalMinutesAsleep ~ SedentaryMinutes, data=daily_data)

summary(Asleep_vs_Sedentary)
```

*4.5 Review the user distribution based on activity level*

Based on a research study by Catrine Tudor-Locke and David Bassett Jr, who propose an index for classification of physical activity based on pedometer measurements, we can, for analysis purposes, establish a classification that fits the four activity levels available in our data.

\< 5,000 steps/day -- Sedentary Lifestyle

5,000 - 7,499 steps/day -- Lightly Active Lifestyle

7,500 - 10,000 steps/day -- Fairly Active Lifestyle

\> 10,000 steps/day -- Very Active Lifestyle

```{r}
#Creating a data frame with average individual data
average_individual_data <- daily_data %>%
  group_by(Id) %>%
  summarise(av_steps=mean(TotalSteps),
            av_calories=mean(Calories),
            av_sleep=mean(TotalMinutesAsleep, na.rm=TRUE))

#Classifying users by level of intensity based on average daily steps
average_individual_data <- average_individual_data %>%
  mutate(user_class=case_when(
    av_steps < 5000 ~ "Sedentary",
    av_steps >= 5000 & av_steps < 7500 ~ "Lightly Active",
    av_steps >= 7500 & av_steps < 10000 ~ "Fairly Active",
    av_steps >= 10000 ~ "Very Active"))

#Checking the percentage of users in each class 
user_activity_class <- average_individual_data %>%
  group_by(user_class) %>%
  summarise(user_count=n()) %>%
  mutate(user_percentage=user_count/sum(user_count)*100)

user_activity_class$user_percentage=paste0(sprintf("%2.1f", user_activity_class$user_percentage), "%")

#Plotting activity class percentage
ggplot(user_activity_class, aes(x="",
                                y=user_percentage,
                                fill=user_class))+
  geom_col()+
  coord_polar("y", start=0)+
  geom_text(aes(label=user_percentage),
            position=position_stack(vjust=0.5),
            size=3)+
  guides(fill=guide_legend(title="User Class"))+
  scale_fill_discrete(breaks=c("Lightly Active",
                               "Fairly Active",
                               "Sedentary",
                               "Very Active"))+
  theme_void()
```

#### Data frame summary

```{r}
project_data_frames <- data.frame(Initial_data_frames=c("daily_activity", "daily_calories", "daily_intensities", "daily_steps", "hourly_calories", "hourly_steps", "sleep", "weight","","",""), Final_data_frames=c("daily_activity", "daily_calories", "daily_intensities", "daily_steps", "hourly_calories", "hourly_steps", "sleep", "average_individual_data", "daily_data", "Intensity_mean", "Intensity_mean_long"))
library(knitr)
kable(project_data_frames, caption="Data frames summary")
```
